 

### 2023-04-24 paper1.Rmd
Revisions in advance of PhD defense, corrections, addition of conceptual framework, additional references

abstract_sentence2 <- paste0("The economic value associated with this burden is estimated at about USD ",
               abstract4,
               " trillion. ")

@@

Evaluating the economic burden of mental illness is a critical part in making the investment case for global mental health, informing public health decision-making, and guiding priority-setting and the scale up of much-needed interventions.[@chisholmScalingupTreatmentDepression2016] At the global level, however, the most recent estimate of the economic impact of mental disorders was published in 2011, using burden of disease estimates from 2004.[@bloomGlobalEconomicBurden2011] This landmark paper by Bloom and colleagues estimated that the value of losses due to mental disorders was roughly 1.3 trillion USD in 2010 (1.6 trillion USD in 2019) when DALYs were valued at one times GDP per capita.[@bloomGlobalEconomicBurden2011] The authors further projected that these losses would grow to nearly 2.5 trillion USD 2010 (or approximately 3.0 trillion USD in 2019) by 2030. (See Tables S1.1 and S1.2 in Appendix 1 for estimates from the other two approaches.) These estimates have been widely cited in calls to action concerning global mental health.[@patelLancetCommissionGlobal2018; @thelancetglobalhealthMentalHealthMatters2020]

@@

The most recent estimates of morbidity and mortality due to mental disorders come from the Global Burden of Disease (GBD) 2019 study.[@vosGlobalBurden3692020a] The GBD study provides disease burden estimates using DALYs, years of life lost (YLLs), and years lived with disability (YLDs), which are then aggregated within a hierarchical grouping scheme that classifies causes of disability and death at different levels of mutually exclusive and completely exhaustive categories. (Mental disorders are a Level 2 condition, nested under noncommunicable diseases (NCDs); see Table S1.3.)

While GBD remains the gold standard for global epidemiologic estimation, the nature of the GBD scheme—in particular, the rationale for grouping certain conditions under mental disorders or not—has been the subject of debate in the literature.[@atunChallengesEstimatingTrue2016; @vigoEstimatingTrueGlobal2016; @whitefordChallengesEstimatingTrue2016] In particular, work by Vigo et al. (2016) published in \emph{The Lancet Psychiatry} argues for an expanded classification of mental disorders under the GBD classification scheme to account for underestimation of the burden of mental disorders.[@vigoEstimatingTrueGlobal2016] The authors attribute this underestimation to five main causes: 1) the distinction drawn between mental and neurological diseases; 2) the categorization of self-harm and suicide under injuries; 3) the classification of all chronic pain and somatoform disorders under musculoskeletal disorders; 4) the exclusion of personality disorders; and 5) the exclusion of premature mortality due to mental disorders. Using data from the 2013 GBD study, Vigo and colleagues re-allocated the entire burden of dementias, epilepsy, migraine, tension-type headache, and self-harm to mental disorders. In addition, a third of the burden of musculoskeletal disorders without anatomical correlate (i.e., somatoform disorders with prominent pain) was attributed to mental disorders.[@vigoEstimatingTrueGlobal2016] This reallocation attributed 13% of DALYs to mental disorders, a 6 percentage point increase from the GBD estimate of 7%.

In this paper, we attempt to revisit the estimation of the global burden of mental disorders and of its associated economic value. Our aim is to characterize potential underestimation of the burden of mental disorders and to quantify the economic value of this burden under different estimation approaches. Specifically, we expand on Vigo et al.[@vigoEstimatingTrueGlobal2016] by capturing premature mortality due to mental disorders using pooled risk ratios of mortality from a systematic review of mental disorders[@walkerMortalityMentalDisorders2015] to determine the population attributable fraction (PAF) of premature mortality. Inclusion of premature mortality through the PAF presents a novel composite approach that can more broadly capture attributable morbidity and mortality. Using this approach on GBD 2019 estimates, we then apply monetary values to DALYs to reach estimates of the global economic value of the mental burden of disease using a value of a statistical life (VSL) approach similar to that of Bloom and colleagues,[@bloomGlobalEconomicBurden2011] which attempts to capture a population’s willingness to pay to reduce morbidity and mortality associated with illness. The VSL approach—in contrast to alternatives such as cost of illness and value of lost output approaches—includes an economic valuation of mortality risk reductions in monetary terms, and thus enables comparison across sectors (beyond the sole health sector) which can motivate decision-making toward ameliorating welfare and societal mental health. Our findings suggest that both the epidemiological and economic burden of mental disorders could be larger than previously estimated, and that underestimation would be larger among regions where premature mortality due to mental disorders is greater.

@@

To estimate the economic burden of mental disorders, we first estimate the attributable mental burden of DALYs under various estimation approaches using data from the 2019 GBD study (available from the Global Health Data Exchange at https://ghdx.healthdata.org/gbd-2019). Second, we apply a monetary value to a DALY to yield an economic assessment associated with these burden estimates.

@@

In our analysis, we replicate the approach of Vigo et al. (2016) using GBD 2019 estimates, applying a similar re-allocation formula to YLLs, YLDs, DALYs, and deaths. Our approach, however, differs in some key respects.

First, we agree with Whiteford and colleagues in viewing the assigning of the entire burden of suicide and self-harm to mental disorders as an overestimate, and consequently do not reallocate all DALYs due to suicide towards the mental health burden.[@whitefordChallengesEstimatingTrue2016] While it is empirically clear that mental disorders elevate the risk of death by suicide and that the majority of suicides appear to be due to mental disorders,[@bertoloteSuicidePsychiatricDiagnosis2002] we view assigning the entirety of this burden to mental disorders as overinclusive, which we avoid to favor a conservative estimation strategy.

@@

Following Vigo and colleagues, we replicate reallocations in neurological and musculoskeletal conditions, and further include alcohol and mental use disorders, as these were previously classified under mental disorders within the GBD classification.

This provides estimates of YLDs due to mental disorders. We then estimate the PAF of mortality due to mental disorders, using GBD prevalence estimates and relative risk estimates for natural-cause and unnatural-cause mortality generated from a systematic review and meta-analysis by Walker et al.[@walkerMortalityMentalDisorders2015] A comparison of our allocation approach with those of Vigo et al. and the original GBD hierarchical allocation is shown in Table 1.1.

@@

Our approach to capturing premature mortality relies on a pooled relative risk estimate for mortality by natural and unnatural causes, drawn from 148 studies identified by Walker et al.[@walkerMortalityMentalDisorders2015] These studies collectively reflect over 338,000 deaths across 29 countries and 6 continents. The majority of deaths (67%) recorded in studies with disaggregated data arose from acute and chronic illnesses, while unnatural causes such as injury and suicide represented 18% of deaths (the rest being unallocated). Overall, the pooled risk of all-cause mortality was 2.2 times higher (95% confidence interval (CI): 2.1-2.3) among people with mental disorders compared to those without. Using this relative risk estimate, Walker and colleagues calculated a PAF to estimate that 8 million deaths were due to mental disorders in 2012.

While Walker and colleagues used a global estimate of the worldwide prevalence of mental disorders in their study to calculate the PAF, we use GBD estimates of prevalence to derive both global- and country-level results. The PAF for a given disorder d and country c is given by:
\[ PAF_{(d,c)}= \frac{p_{(d,c)}\ (RR_d-1)}{1+p_{(d,c)}\ (RR_d-1)}\ \ \ ,\ \ \ \ \ \ \ \ \ \ \ \ (1.1)\]

@@

We separately estimate the PAF for natural and unnatural causes of mortality. Using the calculated PAF estimates, we estimate YLLs attributable to mental disorders by multiplying the PAF by the national burden of mortality. For natural causes of death, we conservatively apply the PAF against YLLs attributable to NCDs. For unnatural causes of death, we apply the PAF against YLLs due to self-harm and injuries. These YLLs are then combined with the YLDs calculated previously to provide DALYs.

@@

To estimate the economic cost associated with premature mortality and morbidity tied to mental illnesses, we assigned a monetary value to attributable DALYs. VSL approaches assign a monetary value to small reductions in mortality risks.[@robinsonValuingMortalityRisk2019] Drawing from these approaches, Jamison and colleagues have estimated monetary values of statistical life years,[@jamisonGlobalHealth20352013] which Khadka and colleagues recently adapted to quantify the economic value of changing mortality risk by cause of death in low- and middle-income countries (LMICs).[@khadkaEconomicValueChanging2021] While VSL approaches are not meant to assign monetary values to full life years or years lived with illness or disability,[@robinsonValuingMortalityRisk2019] the Copenhagen Consensus has previously implemented the use of GDP per capita as a proxy for the monetary value of a DALY as a standard estimate.[@copenhagenconsensuscenterMethodology] Values of one and three times GDP per capita have been suggested as proxies for the value of a DALY.[@bertramCostEffectivenessThresholds2016; @commissiononmacroeconomicsandhealthMacroeconomicsHealthInvesting2001] Estimates of $1,000 and $5,000 per DALY have been used, with the justification that these would be reasonable and convenient lower and upper values, particularly for low-income and lower-middle income countries.[@jamisonInfectiousDiseaseInjury2013; @jhaBenefitsCostsHealth2018]

Consistent with previous approaches, we use GDP per capita (USD 2019) for our base-case value of a DALY. GDP inputs are reported in 2019 USD and obtained from the World Bank’s World Development Indicators; for consistency with our epidemiological inputs, we convert to per capita values using GBD population estimates.

@@

The primary focus of this paper concerns structural uncertainty in determining the burden of mental illness, resulting in the evaluation of three different estimation approaches. To address parameter uncertainty within each approach, we apply a three-way sensitivity analysis. First, following a simple intuitive approach, we incorporate the upper and lower uncertainty intervals (UIs) provided by GBD 2019 for YLLs, YLDs, and DALYs to account for parameter uncertainty. Second, we use the upper and lower values of prevalence estimates and of the 95% confidence intervals (CIs) of the pooled relative risk of all-cause mortality from Walker and colleagues in our composite approach.[@walkerMortalityMentalDisorders2015] Third, our lower bound estimates are set to reallocate one sixth of the burden of musculoskeletal disorders proposed by Vigo and colleagues,[@vigoEstimatingTrueGlobal2016] while our upper bound estimates are set to reallocate one half of this burden.

In addition to our base-case economic valuation, we further report our VSL estimates using three times GDP per capita as the value of a DALY. (Alternative valuations using values of $1,000 and $5,000, as well as purchasing power parity (PPP)-adjusted GDP per capita, are reported in Tables S1.5 and S1.6 of Appendix 1.)

@@

The Institutional Review Board (IRB) of the Harvard T.H. Chan School of Public Health determined that this study was not human subjects research and that additional review was not required (protocol number: IRB20-1946, determined on November 13, 2020).

@@

Rates of DALYs and deaths attributable to mental disorders under the different estimation approaches are presented by GBD region (Figure \ref{figure_gbd_adjusted}). Geographically, the composite approach allocated a large portion of DALYs (to mental disorders) in Eastern Europe, North and Latin America, and sub-Saharan Africa. This is largely driven by the inclusion of premature mortality in the composite approach. Estimates of DALYs by country income group and GBD region are reported in Table 1.2. (Estimates of deaths are reported in Table S1.4-1.6.)

@@

```{r figure_dalys_nation, include = TRUE, echo = FALSE, warnings = FALSE, fig.scap="Disability-adjusted life years (DALYs) attributable to mental disorders in 2019, by estimation approach, as a share of all DALYs.", out.extra='', fig.cap=paste0("\\label{figure_dalys_nation}Disability-adjusted life years (DALYs) attributable to mental disorders in 2019, by estimation approach, as a share of all DALYs. ATG: Antigua and Barbuda. FSM: Federated States of Micronesia. LCA: Saint Lucia. TLS: East Timor. TTO: Trinidad and Tobago. VCT: Saint Vincent and the Grenadines."), fig.width = 13.7, fig.height = 7.5, fig.align='center', out.width = '98%'}

@@

data_temporary <- data_rev_map %>% 
 filter(measure_id == 2) %>% 
 filter(estimate_id == 4) %>%
 st_drop_geometry() %>%
 select(iso_code, percent)

@@

main_map_combined <- cowplot::plot_grid(main_map_upper,
                    main_map_lower,
                    rel_heights = c(5.9, 
                            2.7),
                    rel_widths = c(1.2,1),
                    align='v', 
                    nrow = 2,
                    axis='l')
main_map_combined

@@

ggsave(filename = paste0("Figure_1.png"), 
    plot = last_plot(), 
    path = resultspath)

    # ,
    # width = 1689,
    # height = 1153,
    # units = "px",
    # dpi = 320)

@@

```{r figure_cost_by_region, fig.width = 8, fig.height = 6, fig.align='center', out.width = '95%', fig.scap="Economic burden of mental disorders, as a percent of GDP.", out.extra='', fig.cap="\\label{figure_cost_by_region}Economic burden of mental disorders, as a percent of GDP. The economic value is determined by using GDP per capita (USD 2019) as the value of a DALY. Values are aggregated by GBD region. GDP: gross domestic product. USD: United States dollar. DALY: disability-adjusted life year. GBD: Global Burden of Disease.", echo = FALSE, include = TRUE}

@@

ggsave(filename = paste0("Figure_2.png"),
    path = resultspath,
    width = 8,
    height = 6)

@@

Although economic losses do not represent an actual loss to GDP, a sense of the scale can be gained by expressing the economic consequences with respect to GDP. Figure \ref{figure_cost_by_region} displays the economic burden of disease due to mental disorders under the three estimation approaches by GBD region, as a percent of regional GDP. (Estimates by absolute values per DALY are provided in Appendix 1, along with mapped data visualizing estimates across all values per DALY.) Across approaches, the greatest change in estimated burden occurs in Eastern Europe, Latin America, North America, and Southern sub-Saharan Africa. `r results_sentence3`

@@

Put in context of the existing literature, our epidemiological and economic estimates provide two important contributions. First, our findings echo in magnitude those of Vigo and colleagues,[@vigoEstimatingTrueGlobal2016] which have highlighted that suicide and premature mortality due to mental disorders are potentially large sources of underestimation in the current GBD classification. Second, when including these sources of attributable mortality, the economic findings suggest staggering loses. `r discussion_sentence3`[@bloomGlobalEconomicBurden2011]

Our findings add to a growing literature concerning the classification of mental disorders, in particular related to underscoring the importance of including premature mortality attributable to mental disorders in burden conceptualizations.[@vigoBurdenMentalNeurological2020; @vigoEstimatingTrueGlobal2016; @vigoTrueGlobalDisease2022; @whitefordChallengesEstimatingTrue2016] These calls have most recently been emphasized by GBD collaborators who have urged that “the differential mortality gap for individuals with mental disorders needs to be reflected within the GBD framework.”[@gbd2019mentaldisorderscollaboratorsGlobalRegionalNational2022] Our composite approach to assigning attributable mortality presents one potential attempt for acknowledging this differential mortality gap. Our economic analysis further provides updated monetary estimates of the burden of mental illness; to our knowledge, this is the first such analysis of the global economic burden of mental disorders in over a decade.

Our results should, however, be interpreted with several limitations in mind. First, our estimation approaches themselves all draw upon modeled data (i.e., GBD estimates). While GBD generates descriptions of morbidity and mortality at fine demographic and geographic levels, it is important to emphasize that the sophisticated modeling approaches implemented often draw on (potentially little) available underlying empirical data.[@tichenorMetricPartnershipsGlobal2019] These inputs can be extremely limited for particular diseases and geographical locations, especially so for mental disorders. By way of example, the GBD 2019 Data Input Sources Tool retrieves 3,084 separate data sources for mental disorders. Of these, only 60 pertain to sub-Saharan Africa (1.9%) and 58 to South Asia (1.8%).[@instituteforhealthmetricsandevaluationihmeGlobalBurdenDisease] By comparison, of the 6,064 records pertaining to maternal and neonatal disorders, 631 are for sub-Saharan Africa (10.4%) and 270 for South Asia (4.5%). These severely limited inputs reflect a dearth of global mental health data; as of 2017, the World Mental Health survey initiative had conducted interviews in just 26 countries, only 13 of which were classified as low- or middle-income.[@steinCrossnationalEpidemiologySocial2017]

Relatedly, our composite approach relies on pooled estimates of the relative risk of mortality from a systematic review and meta-analysis that itself is limited by the available data it draws upon.[@walkerMortalityMentalDisorders2015] The review identified 203 studies for inclusion, of which only two were located in Africa, 16 in Asia, and one in South America. While the authors found that the estimates of mortality risk did not vary by region, the limited representation of studies from the world’s most populous and epidemiologically diverse continents is a considerable shortcoming. It is possible, for instance, that the relative risk of all-cause mortality associated with mental disorders is lower where the burden of mortality is more heavily concentrated among child, maternal, and infectious diseases, and is higher where the burden is dominated by NCDs. Therefore, to reach a conservative estimate of attributable mortality, we separately estimated population attributable fractions for natural and unnatural causes of death and restricted our allocation of YLLs from natural causes to NCDs—meaning no deaths from maternal or infectious diseases were attributed to mental disorders under the composite approach.

Furthermore, our composite approach allocates mortality due to mental disorders by calculating population attributable fractions using the conventional formula, which may be biased in the presence of confounding or effect heterogeneity.[@darrowConfoundingBiasAttributable2011] In particular, the use of adjusted risk ratios (as in the current analysis) may result in anticonservative bias if the crude risk ratios are lower than the adjusted ones. To mitigate the potential for bias, our sensitivity analysis presents results under conservative assumptions for risk ratios and estimates of prevalence and mortality.

>>>>

abstract_sentence2 <- paste0("The economic value associated with this burden is estimated at about ",
               abstract4,
               " trillion United States dollars. ")

@@

Evaluating the economic burden of mental illness is a critical part in making the investment case for global mental health, informing public health decision-making, and guiding priority-setting and the scale up of much-needed interventions.[@chisholmScalingupTreatmentDepression2016] At the global level, however, the most recent estimate of the economic impact of mental disorders was published in 2011, using burden of disease estimates from 2004.[@bloomGlobalEconomicBurden2011] This landmark paper by Bloom and colleagues estimated that the value of losses due to mental disorders was roughly 1.3 trillion United States dollars (\acr{USD}) in 2010 (1.6 trillion \acr{USD} in 2019) when disability-adjusted life years (\acr{DALYs}) were valued at one times \acr{GDP} per capita.[@bloomGlobalEconomicBurden2011] The authors further projected that these losses would grow to nearly 2.5 trillion \acr{USD} 2010 (or approximately 3.0 trillion \acr{USD} in 2019) by 2030. (See Tables S1.1 and S1.2 in Appendix 1 for estimates from the other two approaches.) These estimates have been widely cited in calls to action concerning global mental health.[@patelLancetCommissionGlobal2018; @thelancetglobalhealthMentalHealthMatters2020]

@@

The most recent estimates of morbidity and mortality due to mental disorders come from the Global Burden of Disease (\acr{GBD}) 2019 study.[@vosGlobalBurden3692020a] The \acr{GBD} study provides disease burden estimates using \acr{DALYs}, years of healthy life lost due to disability (\acr{YLDs}), and \acr{YLLs}, which are then aggregated within a hierarchical grouping scheme that classifies causes of disability and death at different levels of mutually exclusive and completely exhaustive categories. (Mental disorders are a Level 2 condition, nested under \acr{NCDs}; see Table S1.3.)

While \acr{GBD} remains the gold standard for global epidemiologic estimation, the nature of the \acr{GBD} scheme—in particular, the rationale for grouping certain conditions under mental disorders or not—has been the subject of debate in the literature.[@atunChallengesEstimatingTrue2016; @vigoEstimatingTrueGlobal2016; @whitefordChallengesEstimatingTrue2016] In particular, work by Vigo et al. (2016) published in \emph{The Lancet Psychiatry} argues for an expanded classification of mental disorders under the \acr{GBD} classification scheme to account for underestimation of the burden of mental disorders.[@vigoEstimatingTrueGlobal2016] The authors attribute this underestimation to five main causes: 1) the distinction drawn between mental and neurological diseases; 2) the categorization of self-harm and suicide under injuries; 3) the classification of all chronic pain and somatoform disorders under musculoskeletal disorders; 4) the exclusion of personality disorders; and 5) the exclusion of premature mortality due to mental disorders. Using data from the 2013 \acr{GBD} study, Vigo and colleagues re-allocated the entire burden of dementias, epilepsy, migraine, tension-type headache, and self-harm to mental disorders. In addition, a third of the burden of musculoskeletal disorders without anatomical correlate (i.e., somatoform disorders with prominent pain) was attributed to mental disorders.[@vigoEstimatingTrueGlobal2016] This reallocation attributed 13% of \acr{DALYs} to mental disorders, a 6 percentage point increase from the \acr{GBD} estimate of 7%.

In this paper, we attempt to revisit the estimation of the global burden of mental disorders and of its associated economic value. Our aim is to characterize potential underestimation of the burden of mental disorders and to quantify the economic value of this burden under different estimation approaches. Specifically, we expand on Vigo et al.[@vigoEstimatingTrueGlobal2016] by capturing premature mortality due to mental disorders using pooled risk ratios of mortality from a systematic review of mental disorders[@walkerMortalityMentalDisorders2015] to determine the \acr{PAF} of premature mortality. Inclusion of premature mortality through the \acr{PAF} presents a novel composite approach that can more broadly capture attributable morbidity and mortality. Using this approach on \acr{GBD} 2019 estimates, we then apply monetary values to \acr{DALYs} to reach estimates of the global economic value of the mental burden of disease using a \acr{VSL} approach similar to that of Bloom and colleagues,[@bloomGlobalEconomicBurden2011] which attempts to capture a population’s willingness to pay to reduce morbidity and mortality associated with illness. The \acr{VSL} approach—in contrast to alternatives such as cost of illness and value of lost output approaches—includes an economic valuation of mortality risk reductions in monetary terms, and thus enables comparison across sectors (beyond the sole health sector) which can motivate decision-making toward ameliorating welfare and societal mental health. Our findings suggest that both the epidemiological and economic burden of mental disorders could be larger than previously estimated, and that underestimation would be larger among regions where premature mortality due to mental disorders is greater.

@@

To estimate the economic burden of mental disorders, we first estimate the attributable mental burden of \acr{DALYs} under various estimation approaches using data from the 2019 \acr{GBD} study (available from the Global Health Data Exchange at https://ghdx.healthdata.org/gbd-2019). Second, we apply a monetary value to a DALY to yield an economic assessment associated with these burden estimates.

@@

In our analysis, we replicate the approach of Vigo et al. (2016) using \acr{GBD} 2019 estimates, applying a similar re-allocation formula to YLLs, YLDs, \acr{DALYs}, and deaths. Our approach, however, differs in some key respects.

First, we agree with Whiteford and colleagues in viewing the assigning of the entire burden of suicide and self-harm to mental disorders as an overestimate, and consequently do not reallocate all \acr{DALYs} due to suicide towards the mental health burden.[@whitefordChallengesEstimatingTrue2016] While it is empirically clear that mental disorders elevate the risk of death by suicide and that the majority of suicides appear to be due to mental disorders,[@bertoloteSuicidePsychiatricDiagnosis2002] we view assigning the entirety of this burden to mental disorders as overinclusive, which we avoid to favor a conservative estimation strategy.

@@

Following Vigo and colleagues, we replicate reallocations in neurological and musculoskeletal conditions, and further include alcohol and mental use disorders, as these were previously classified under mental disorders within the \acr{GBD} classification.

This provides estimates of YLDs due to mental disorders. We then estimate the \acr{PAF} of mortality due to mental disorders, using \acr{GBD} prevalence estimates and relative risk estimates for natural-cause and unnatural-cause mortality generated from a systematic review and meta-analysis by Walker et al.[@walkerMortalityMentalDisorders2015] A comparison of our allocation approach with those of Vigo et al. and the original \acr{GBD} hierarchical allocation is shown in Table 1.1.

@@

Our approach to capturing premature mortality relies on a pooled relative risk estimate for mortality by natural and unnatural causes, drawn from 148 studies identified by Walker et al.[@walkerMortalityMentalDisorders2015] These studies collectively reflect over 338,000 deaths across 29 countries and 6 continents. The majority of deaths (67%) recorded in studies with disaggregated data arose from acute and chronic illnesses, while unnatural causes such as injury and suicide represented 18% of deaths (the rest being unallocated). Overall, the pooled risk of all-cause mortality was 2.2 times higher (95% confidence interval (CI): 2.1-2.3) among people with mental disorders compared to those without. Using this relative risk estimate, Walker and colleagues calculated a \acr{PAF} to estimate that 8 million deaths were due to mental disorders in 2012.

While Walker and colleagues used a global estimate of the worldwide prevalence of mental disorders in their study to calculate the PAF, we use \acr{GBD} estimates of prevalence to derive both global- and country-level results. The \acr{PAF} for a given disorder d and country c is given by:
\[ PAF_{(d,c)}= \frac{p_{(d,c)}\ (RR_d-1)}{1+p_{(d,c)}\ (RR_d-1)}\ \ \ ,\ \ \ \ \ \ \ \ \ \ \ \ (1.1)\]

@@

We separately estimate the \acr{PAF} for natural and unnatural causes of mortality. Using the calculated \acr{PAF} estimates, we estimate YLLs attributable to mental disorders by multiplying the \acr{PAF} by the national burden of mortality. For natural causes of death, we conservatively apply the \acr{PAF} against YLLs attributable to \acr{NCDs}. For unnatural causes of death, we apply the \acr{PAF} against YLLs due to self-harm and injuries. These YLLs are then combined with the YLDs calculated previously to provide \acr{DALYs}.

@@

To estimate the economic cost associated with premature mortality and morbidity tied to mental illnesses, we assigned a monetary value to attributable \acr{DALYs}. \acr{VSL} approaches assign a monetary value to small reductions in mortality risks.[@robinsonValuingMortalityRisk2019] Drawing from these approaches, Jamison and colleagues have estimated monetary values of statistical life years,[@jamisonGlobalHealth20352013] which Khadka and colleagues recently adapted to quantify the economic value of changing mortality risk by cause of death in \acr{LMICs}.[@khadkaEconomicValueChanging2021] While \acr{VSL} approaches are not meant to assign monetary values to full life years or years lived with illness or disability,[@robinsonValuingMortalityRisk2019] the Copenhagen Consensus has previously implemented the use of \acr{GDP} per capita as a proxy for the monetary value of a DALY as a standard estimate.[@copenhagenconsensuscenterMethodology] Values of one and three times \acr{GDP} per capita have been suggested as proxies for the value of a DALY.[@bertramCostEffectivenessThresholds2016; @commissiononmacroeconomicsandhealthMacroeconomicsHealthInvesting2001] Estimates of $1,000 and $5,000 per DALY have been used, with the justification that these would be reasonable and convenient lower and upper values, particularly for low-income and lower-middle income countries.[@jamisonInfectiousDiseaseInjury2013; @jhaBenefitsCostsHealth2018]

Consistent with previous approaches, we use \acr{GDP} per capita (\acr{USD} 2019) for our base-case value of a DALY. \acr{GDP} inputs are reported in 2019 \acr{USD} and obtained from the World Bank’s World Development Indicators; for consistency with our epidemiological inputs, we convert to per capita values using \acr{GBD} population estimates.

@@

The primary focus of this paper concerns structural uncertainty in determining the burden of mental illness, resulting in the evaluation of three different estimation approaches. To address parameter uncertainty within each approach, we apply a three-way sensitivity analysis. First, following a simple intuitive approach, we incorporate the upper and lower \acr{UIs} provided by \acr{GBD} 2019 for \acr{YLLs}, \acr{YLDs}, and \acr{DALYs} to account for parameter uncertainty. Second, we use the upper and lower values of prevalence estimates and of the 95% \acr{CIs} of the pooled relative risk of all-cause mortality from Walker and colleagues in our composite approach.[@walkerMortalityMentalDisorders2015] Third, our lower bound estimates are set to reallocate one sixth of the burden of musculoskeletal disorders proposed by Vigo and colleagues,[@vigoEstimatingTrueGlobal2016] while our upper bound estimates are set to reallocate one half of this burden.

In addition to our base-case economic valuation, we further report our \acr{VSL} estimates using three times \acr{GDP} per capita as the value of a DALY. (Alternative valuations using values of $1,000 and $5,000, as well as \acr{PPP}-adjusted \acr{GDP} per capita, are reported in Tables S1.5 and S1.6 of Appendix 1.)

@@

The \acr{IRB} of the Harvard T.H. Chan School of Public Health determined that this study was not human subjects research and that additional review was not required (protocol number: IRB20-1946, determined on November 13, 2020).

@@

Rates of \acr{DALYs} and deaths attributable to mental disorders under the different estimation approaches are presented by \acr{GBD} region (Figure \ref{figure_gbd_adjusted}). Geographically, the composite approach allocated a large portion of \acr{DALYs} (to mental disorders) in Eastern Europe, North and Latin America, and sub-Saharan Africa. This is largely driven by the inclusion of premature mortality in the composite approach. Estimates of DALYs by country income group and \acr{GBD} region are reported in Table 1.2. (Estimates of deaths are reported in Table S1.4-1.6.)

@@

```{r figure_dalys_nation, include = TRUE, echo = FALSE, warnings = FALSE, fig.scap="Disability-adjusted life years (DALYs) attributable to mental disorders in 2019, under the composite estimation approach, as a share of all DALYs.", out.extra='', fig.cap=paste0("\\label{figure_dalys_nation}Disability-adjusted life years (DALYs) attributable to mental disorders in 2019, under the composite estimation approach, as a share of all DALYs. ATG: Antigua and Barbuda. FSM: Federated States of Micronesia. LCA: Saint Lucia. TLS: East Timor. TTO: Trinidad and Tobago. VCT: Saint Vincent and the Grenadines."), fig.width = 13.7, fig.height = 7.5, fig.align='center', out.width = '98%'}

@@

estimates_values <- c(1,2,4)

for (z in estimates_values) {

data_temporary <- data_rev_map %>% 
 filter(measure_id == 2) %>% 
 filter(estimate_id == z) %>%
 st_drop_geometry() %>%
 select(iso_code, percent)

@@

main_map_combined <- cowplot::plot_grid(main_map_upper,
                    main_map_lower,
                    rel_heights = c(5.9, 
                            2.7),
                    rel_widths = c(1.2,1),
                    align='v', 
                    nrow = 2,
                    axis='l')

ggsave(filename = paste0("Figure_2_estimate_",z,".jpg"), 
    plot = final_figure, 
    width = 13.33,
    height = 6.3,
    bg = "white")

if (z == 4) {

  main_map_combined

@@

ggsave(filename = paste0("Figure_3_DALY.png"), 
    plot = dalys, 
    width = 12,
    height = 6.3,
    bg = "white")

ggsave(filename = paste0("Figure_3_YLD.png"), 
    plot = ylds, 
    width = 12,
    height = 6.3,
    bg = "white")

ggsave(filename = paste0("Figure_3_YLL.png"), 
    plot = ylls, 
    width = 12,
    height = 6.3,
    bg = "white")

ggsave(filename = paste0("Figure_3_deaths.png"), 
    plot = deaths, 
    width = 12,
    height = 6.3,
    bg = "white")

@@

```{r figure_cost_by_region, fig.width = 8, fig.height = 6, fig.align='center', out.width = '95%', fig.scap="Economic burden of mental disorders, as a percent of GDP.", out.extra='', fig.cap="\\label{figure_cost_by_region}Economic burden of mental disorders, as a percent of GDP. The economic value is determined by using GDP per capita (USD 2019) as the value of a DALY. Values are aggregated by GBD region. GDP: gross domestic product. USD: United States dollars. DALY: disability-adjusted life year. GBD: Global Burden of Disease.", echo = FALSE, include = TRUE}

@@

ggsave(filename = paste0("Figure_4.png"), 
    plot = last_plot(), 
    width = 8,
    height = 6,
    bg = "white")

@@

Although economic losses do not represent an actual loss to \acr{GDP}, a sense of the scale can be gained by expressing the economic consequences with respect to \acr{GDP}. Figure \ref{figure_cost_by_region} displays the economic burden of disease due to mental disorders under the three estimation approaches by \acr{GBD} region, as a percent of regional \acr{GDP}. (Estimates by absolute values per DALY are provided in Appendix 1, along with mapped data visualizing estimates across all values per DALY.) Across approaches, the greatest change in estimated burden occurs in Eastern Europe, Latin America, North America, and Southern sub-Saharan Africa. `r results_sentence3`

@@

Put in context of the existing literature, our epidemiological and economic estimates provide two important contributions. First, our findings echo in magnitude those of Vigo and colleagues,[@vigoEstimatingTrueGlobal2016] which have highlighted that suicide and premature mortality due to mental disorders are potentially large sources of underestimation in the current \acr{GBD} classification. Second, when including these sources of attributable mortality, the economic findings suggest staggering loses. `r discussion_sentence3`[@bloomGlobalEconomicBurden2011]

Our findings add to a growing literature concerning the classification of mental disorders, in particular related to underscoring the importance of including premature mortality attributable to mental disorders in burden conceptualizations.[@vigoBurdenMentalNeurological2020; @vigoEstimatingTrueGlobal2016; @vigoTrueGlobalDisease2022; @whitefordChallengesEstimatingTrue2016] These calls have most recently been emphasized by \acr{GBD} collaborators who have urged that “the differential mortality gap for individuals with mental disorders needs to be reflected within the \acr{GBD} framework.”[@gbd2019mentaldisorderscollaboratorsGlobalRegionalNational2022] Our composite approach to assigning attributable mortality presents one potential attempt for acknowledging this differential mortality gap. Our economic analysis further provides updated monetary estimates of the burden of mental illness; to our knowledge, this is the first such analysis of the global economic burden of mental disorders in over a decade.

Our results should, however, be interpreted with several limitations in mind. First, our estimation approaches themselves all draw upon modeled data (i.e., \acr{GBD} estimates). While \acr{GBD} generates descriptions of morbidity and mortality at fine demographic and geographic levels, it is important to emphasize that the sophisticated modeling approaches implemented often draw on (potentially little) available underlying empirical data.[@tichenorMetricPartnershipsGlobal2019] These inputs can be extremely limited for particular diseases and geographical locations, especially so for mental disorders. By way of example, the \acr{GBD} 2019 Data Input Sources Tool retrieves 3,084 separate data sources for mental disorders. Of these, only 60 pertain to sub-Saharan Africa (1.9%) and 58 to South Asia (1.8%).[@instituteforhealthmetricsandevaluationihmeGlobalBurdenDisease] By comparison, of the 6,064 records pertaining to maternal and neonatal disorders, 631 are for sub-Saharan Africa (10.4%) and 270 for South Asia (4.5%). These severely limited inputs reflect a dearth of global mental health data; as of 2017, the World Mental Health survey initiative had conducted interviews in just 26 countries, only 13 of which were classified as low- or middle-income.[@steinCrossnationalEpidemiologySocial2017]

Relatedly, our composite approach relies on pooled estimates of the relative risk of mortality from a systematic review and meta-analysis that itself is limited by the available data it draws upon.[@walkerMortalityMentalDisorders2015] The review identified 203 studies for inclusion, of which only two were located in Africa, 16 in Asia, and one in South America. While the authors found that the estimates of mortality risk did not vary by region, the limited representation of studies from the world’s most populous and epidemiologically diverse continents is a considerable shortcoming. It is possible, for instance, that the relative risk of all-cause mortality associated with mental disorders is lower where the burden of mortality is more heavily concentrated among child, maternal, and infectious diseases, and is higher where the burden is dominated by \acr{NCDs}. Therefore, to reach a conservative estimate of attributable mortality, we separately estimated \acr{PAF}s for natural and unnatural causes of death and restricted our allocation of YLLs from natural causes to \acr{NCDs}—meaning no deaths from maternal or infectious diseases were attributed to mental disorders under the composite approach.

Furthermore, our composite approach allocates mortality due to mental disorders by calculating \acr{PAF}s using the conventional formula, which may be biased in the presence of confounding or effect heterogeneity.[@darrowConfoundingBiasAttributable2011] In particular, the use of adjusted risk ratios (as in the current analysis) may result in anticonservative bias if the crude risk ratios are lower than the adjusted ones. To mitigate the potential for bias, our sensitivity analysis presents results under conservative assumptions for risk ratios and estimates of prevalence and mortality.
